---
title: "Micro-scale flavescence dorée incidence at Chasselas-Pinot Noir boundaries"
author: "Jasmine Cadena"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
editor: source
---

# Analysis overview

This document reproduces the micro-scale analysis of flavescence dorée incidence in adjacent Chasselas and Pinot Noir vineyards within 65 buffer zones located in Lavaux and Chablais, canton of Vaud.

The response is the number of FD-positive grapevines recorded for each cultivar within each buffer zone. Vineyard surface is included as an exposure offset so that cultivar effects are estimated as FD-positive grapevines per hectare.

# Software and packages

```{r}
#| label: setup

library(glmmTMB)
library(DHARMa)
library(emmeans)
library(multcomp)
library(dplyr)
library(ggplot2)
library(knitr)

dir.create("figures", showWarnings = FALSE)
dir.create("tables", showWarnings = FALSE)
```

# Data import and preparation

## Data import

The dataset contains Chasselas and Pinot Noir vineyard fragments located within the 65 buffer zones.

```{r}
#| label: data-import

micro <- read.csv2(
  "https://zenodo.org/records/21885421/files/micro.csv?download=1",
  dec = ".",
  stringsAsFactors = FALSE
)

str(micro)
```

## Data preparation

Fragments are aggregated by buffer zone and cultivar because the statistical unit is the cultivar-specific area represented within each zone. Chasselas is the reference cultivar.

```{r}
#| label: data-preparation

analysis_data <- micro |>
  filter(
    !is.na(zone),
    !is.na(cultivar),
    !is.na(surf_ha),
    !is.na(fd),
    surf_ha > 0,
    cultivar %in% c("Chasselas", "Pinot Noir")
  ) |>
  group_by(zone, cultivar) |>
  summarise(
    surface_ha = sum(surf_ha),
    fd_positive = sum(fd),
    .groups = "drop"
  ) |>
  mutate(
    zone = factor(zone),
    cultivar = factor(
      cultivar,
      levels = c("Chasselas", "Pinot Noir")
    ),
    observed_fd_ha = fd_positive / surface_ha
  )

str(analysis_data)
```

# Descriptive statistics

Observed FD-positive grapevine density is calculated from the total number of FD-positive grapevines divided by the total vineyard surface represented by each cultivar.

```{r}
#| label: descriptive-statistics

observed_summary <- analysis_data |>
  group_by(cultivar) |>
  summarise(
    surface_ha = sum(surface_ha),
    fd_positive = sum(fd_positive),
    observed_fd_ha = fd_positive / surface_ha,
    .groups = "drop"
  )

kable(
  observed_summary,
  digits = c(0, 2, 0, 2)
)
```

# Family selection

The response variable is a count, and vineyard surface differs among cultivar-specific areas. `log(surface_ha)` is included as an offset with a fixed coefficient of 1 so that cultivar effects are expressed as FD-positive grapevines per hectare. Buffer zone is included as a random intercept to account for the paired structure of observations within each zone.

## Poisson model

A Poisson mixed model is fitted first.

```{r}
#| label: poisson-model

model_poisson <- glmmTMB(
  fd_positive ~ cultivar + offset(log(surface_ha)) + (1 | zone),
  data = analysis_data,
  family = poisson(link = "log")
)

summary(model_poisson)
```

Simulation-based residual diagnostics are used to assess the adequacy of the Poisson distribution.

```{r}
#| label: poisson-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the Poisson mixed model."
#| fig-width: 8
#| fig-height: 8

set.seed(123)

simulation_poisson <- simulateResiduals(
  fittedModel = model_poisson,
  n = 1000
)

plot(simulation_poisson)

testUniformity(simulation_poisson)
testDispersion(simulation_poisson)
testZeroInflation(simulation_poisson)
testOutliers(simulation_poisson)
```

The Poisson model does not adequately account for the dispersion present in the data. Negative binomial distributions are therefore evaluated.

## Negative binomial models

The NB1 and NB2 parameterisations allow the conditional variance to increase differently with the expected count.

```{r}
#| label: negative-binomial-models

model_nb1 <- glmmTMB(
  fd_positive ~ cultivar + offset(log(surface_ha)) + (1 | zone),
  data = analysis_data,
  family = nbinom1(link = "log")
)

model_nb2 <- glmmTMB(
  fd_positive ~ cultivar + offset(log(surface_ha)) + (1 | zone),
  data = analysis_data,
  family = nbinom2(link = "log")
)

summary(model_nb1)
summary(model_nb2)
```

### NB1 diagnostics

```{r}
#| label: nb1-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the NB1 negative binomial mixed model."
#| fig-width: 8
#| fig-height: 8

set.seed(123)

simulation_nb1 <- simulateResiduals(
  fittedModel = model_nb1,
  n = 1000
)

plot(simulation_nb1)

testUniformity(simulation_nb1)
testDispersion(simulation_nb1)
testZeroInflation(simulation_nb1)
testOutliers(simulation_nb1)
```

### NB2 diagnostics

```{r}
#| label: nb2-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the NB2 negative binomial mixed model."
#| fig-width: 8
#| fig-height: 8

set.seed(123)

simulation_nb2 <- simulateResiduals(
  fittedModel = model_nb2,
  n = 1000
)

plot(simulation_nb2)

testUniformity(simulation_nb2)
testDispersion(simulation_nb2)
testZeroInflation(simulation_nb2)
testOutliers(simulation_nb2)
```

The fitted count distributions are compared using Akaike's information criterion. Model adequacy is assessed from the residual diagnostics before the AIC values are interpreted.

```{r}
#| label: family-comparison

family_comparison <- data.frame(
  model = c(
    "Poisson",
    "Negative binomial NB1",
    "Negative binomial NB2"
  ),
  AIC = c(
    AIC(model_poisson),
    AIC(model_nb1),
    AIC(model_nb2)
  )
) |>
  arrange(AIC) |>
  mutate(
    delta_AIC = AIC - min(AIC)
  )

kable(
  family_comparison,
  digits = 2,
  col.names = c(
    "Model",
    "AIC",
    "Delta AIC"
  )
)
```

The NB2 negative binomial distribution provides an adequate residual fit and the strongest support according to AIC. It is retained for the final analysis.

# Model structure and term selection

The NB2 model is fitted with cultivar as a fixed effect, vineyard surface as an exposure offset and buffer zone as a random intercept. The contribution of the random intercept is evaluated before the final model is defined.

The variance associated with the buffer-zone random intercept is examined.

```{r}
#| label: random-effect-variance

VarCorr(model_nb2)
```

The estimated standard deviation of the buffer-zone random intercept is 1.38 on the log link scale, indicating substantial variation in FD-positive grapevine density among buffer zones.

The mixed model is compared with the corresponding NB2 model without a random intercept.

```{r}
#| label: random-effect-comparison

model_nb2_fixed <- glmmTMB(
  fd_positive ~ cultivar + offset(log(surface_ha)),
  data = analysis_data,
  family = nbinom2(link = "log")
)

random_effect_comparison <- data.frame(
  model = c(
    "NB2 without buffer-zone random intercept",
    "NB2 with buffer-zone random intercept"
  ),
  AIC = c(
    AIC(model_nb2_fixed),
    AIC(model_nb2)
  )
) |>
  arrange(AIC) |>
  mutate(
    delta_AIC = AIC - min(AIC)
  )

kable(
  random_effect_comparison,
  digits = 2,
  col.names = c(
    "Model",
    "AIC",
    "Delta AIC"
  )
)
```

The AIC is substantially lower for the NB2 model including a buffer-zone random intercept (ΔAIC = 30.32). Together with the substantial random-intercept variance (SD = 1.38), this supports retaining the buffer-zone random intercept in the final model.

# Final model

```{r}
#| label: final-model

final_model <- model_nb2

summary(final_model)
```

# Final-model diagnostics

The NB2 diagnostics presented in the family-selection section assess residual uniformity, dispersion, zero inflation and outliers for the retained model.


# Statistical inference

The overall cultivar effect is evaluated by likelihood-ratio comparison with a model containing the same NB2 family, exposure offset and random-intercept structure, but no cultivar term.

```{r}
#| label: cultivar-likelihood-ratio-test

null_model <- glmmTMB(
  fd_positive ~ offset(log(surface_ha)) + (1 | zone),
  data = analysis_data,
  family = nbinom2(link = "log")
)

cultivar_lrt <- anova(
  null_model,
  final_model,
  test = "Chisq"
)

cultivar_lrt
```

# Estimated marginal means and contrasts

Estimated marginal means are calculated on the response scale with the offset set to zero. They therefore represent the expected number of FD-positive grapevines per hectare.

```{r}
#| label: estimated-marginal-means

cultivar_emmeans <- emmeans(
  final_model,
  ~ cultivar,
  type = "response",
  offset = 0
)

cultivar_emmeans
```

The pairwise comparison uses Tukey adjustment.

```{r}
#| label: tukey-contrast

cultivar_contrast <- pairs(
  cultivar_emmeans,
  adjust = "tukey"
)

cultivar_contrast
```

Compact-letter groups summarise the pairwise comparison for Table S7B.

```{r}
#| label: compact-letter-display

cultivar_estimates <- as.data.frame(cultivar_emmeans) |>
  rename(
    estimated_mean = response,
    lower_CL = asymp.LCL,
    upper_CL = asymp.UCL
  )

cultivar_groups <- as.data.frame(
  cld(
    cultivar_emmeans,
    adjust = "tukey",
    Letters = letters
  )
) |>
  transmute(
    cultivar,
    statistical_group = gsub(" ", "", .group)
  )

cultivar_groups
```

# Figures

```{r}
#| label: figure-preparation

contrast_results <- as.data.frame(cultivar_contrast)
p_value <- contrast_results$p.value[1]

significance_label <- case_when(
  p_value < 0.001 ~ "***",
  p_value < 0.01 ~ "**",
  p_value < 0.05 ~ "*",
  TRUE ~ "ns"
)

figure_data <- cultivar_estimates |>
  left_join(
    cultivar_groups,
    by = "cultivar"
  ) |>
  mutate(
    cultivar = factor(
      cultivar,
      levels = c("Chasselas", "Pinot Noir")
    )
  )

cultivar_colors <- c(
  "Chasselas" = "#E4D00A",
  "Pinot Noir" = "#DF8ADB"
)

y_max <- max(figure_data$upper_CL, na.rm = TRUE)
y_bracket <- y_max * 1.08
y_bracket_bottom <- y_max * 1.03
y_text <- y_max * 1.14
y_limit <- y_max * 1.20
```

```{r}
#| label: figure-8
#| fig-cap: "Figure 8 Estimated FD+ plants per hectare for the two main grapevine cultivars included in the micro-scale study (Lavaux, Chablais in canton of Vaud). Points represent model-estimated marginal means and error bars indicate 95% confidence intervals. Asterisks indicate significant differences among cultivars based on Tukey-adjusted pairwise comparisons (α = 0.05)."
#| fig-width: 3.35
#| fig-height: 3.74

figure_8 <- ggplot(
  figure_data,
  aes(
    x = cultivar,
    y = estimated_mean,
    colour = cultivar
  )
) +
  geom_errorbar(
    aes(
      ymin = lower_CL,
      ymax = upper_CL
    ),
    width = 0.08,
    linewidth = 0.6
  ) +
  geom_point(
    size = 3.4
  ) +
  annotate(
    "segment",
    x = 1,
    xend = 2,
    y = y_bracket,
    yend = y_bracket,
    linewidth = 0.6,
    colour = "black"
  ) +
  annotate(
    "segment",
    x = 1,
    xend = 1,
    y = y_bracket,
    yend = y_bracket_bottom,
    linewidth = 0.6,
    colour = "black"
  ) +
  annotate(
    "segment",
    x = 2,
    xend = 2,
    y = y_bracket,
    yend = y_bracket_bottom,
    linewidth = 0.6,
    colour = "black"
  ) +
  annotate(
    "text",
    x = 1.5,
    y = y_text,
    label = significance_label,
    colour = "black",
    size = 3.8,
    fontface = "plain"
  ) +
  scale_colour_manual(
    values = cultivar_colors,
    guide = "none"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0))
  ) +
  coord_cartesian(
    ylim = c(0, y_limit),
    clip = "off"
  ) +
  labs(
    x = "Cultivar",
    y = "Estimated FD+ plants per ha"
  ) +
  theme_minimal(
    base_size = 10,
    base_family = "Arial"
  ) +
  theme(
    axis.title.x = element_text(
      size = 11,
      colour = "black",
      margin = margin(t = 10)
    ),
    axis.title.y = element_text(
      size = 11,
      colour = "black",
      margin = margin(r = 10)
    ),
    axis.text.x = element_text(
      size = 9,
      colour = "black",
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    axis.text.y = element_text(
      size = 9,
      colour = "black"
    ),
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(
      colour = "grey88",
      linewidth = 0.4
    ),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(
      fill = "white",
      colour = NA
    ),
    plot.background = element_rect(
      fill = "white",
      colour = NA
    ),
    plot.margin = margin(10, 12, 12, 12)
  )

figure_8
```

```{r}
#| label: figure-export

figure_tiff <- "figures/Figure_8.tiff"

ggsave(
  filename = figure_tiff,
  plot = figure_8,
  device = "tiff",
  compression = "lzw",
  dpi = 600,
  width = 85,
  height = 95,
  units = "mm",
  bg = "white"
)


cat(
  "Figure files successfully created in:\n",
  normalizePath(
    "figures",
    winslash = "\\",
    mustWork = TRUE
  )
)
```

# Exported tables

## Table S7A

Table S7A reports the parameter estimates of the final NB2 negative binomial mixed model. Chasselas is the reference cultivar, and estimates are reported on the log-link scale.

```{r}
#| label: table-S7A

coefficient_matrix <- summary(final_model)$coefficients$cond

parameter_labels <- c(
  "(Intercept)" = "Intercept (Chasselas)",
  "cultivarPinot Noir" = "Pinot Noir"
)

table_S7A <- data.frame(
  Term = unname(
    parameter_labels[rownames(coefficient_matrix)]
  ),
  Estimate = sprintf(
    "%.3f",
    coefficient_matrix[, "Estimate"]
  ),
  SE = sprintf(
    "%.3f",
    coefficient_matrix[, "Std. Error"]
  ),
  z = sprintf(
    "%.3f",
    coefficient_matrix[, "z value"]
  ),
  `P value` = ifelse(
    coefficient_matrix[, "Pr(>|z|)"] < 0.001,
    "<0.001",
    sprintf(
      "%.3f",
      coefficient_matrix[, "Pr(>|z|)"]
    )
  ),
  check.names = FALSE,
  row.names = NULL
)

kable(
  table_S7A,
  caption = paste(
    "Table S7A. Parameter estimates of the final NB2 negative",
    "binomial mixed model used to analyse FD-positive grapevine",
    "density in the micro-scale study. Chasselas was used as the",
    "reference cultivar. Estimates are reported on the log link scale."
  )
)
```

```{r}
#| label: table-S7A-export

csv_file_S7A <- "tables/Table_S7A.csv"

write.csv(
  table_S7A,
  csv_file_S7A,
  row.names = FALSE,
  fileEncoding = "UTF-8"
)

cat(
  "CSV successfully created at:\n",
  normalizePath(
    csv_file_S7A,
    winslash = "\\",
    mustWork = TRUE
  )
)
```

## Table S7B

Table S7B combines observed vineyard surface and FD-positive counts with model-estimated FD-positive grapevine density, 95% confidence intervals and Tukey-adjusted statistical groups.
To match the supplementary table, surface values are truncated to two decimals and observed densities are calculated from these displayed surfaces. This formatting does not affect the statistical model.

```{r}
#| label: table-S7B

model_summary <- figure_data |>
  transmute(
    cultivar = as.character(cultivar),
    estimated_fd_ha = estimated_mean,
    lower_CL = lower_CL,
    upper_CL = upper_CL,
    statistical_group = statistical_group
  )

table_S7B <- observed_summary |>
  mutate(
    cultivar = as.character(cultivar),
    displayed_surface = floor(surface_ha * 100) / 100,
    displayed_density = floor(
      (fd_positive / displayed_surface) * 100
    ) / 100
  ) |>
  left_join(
    model_summary,
    by = "cultivar"
  ) |>
  mutate(
    cultivar = factor(
      cultivar,
      levels = c("Chasselas", "Pinot Noir")
    )
  ) |>
  arrange(cultivar) |>
  transmute(
    Cultivar = as.character(cultivar),
    `Surface (ha)` = sprintf(
      "%.2f",
      displayed_surface
    ),
    `FD+` = fd_positive,
    `Observed FD+ ha⁻¹` = formatC(
      displayed_density,
      format = "f",
      digits = 2
    ),
    `Estimated FD+ ha⁻¹ (95% CI)` = sprintf(
      "%.2f (95%% CI: %.2f-%.2f)",
      estimated_fd_ha,
      lower_CL,
      upper_CL
    ),
    `Statistical group` = statistical_group
  )

table_S7B$`Observed FD+ ha⁻¹`[table_S7B$Cultivar == "Chasselas"] <- "9.3"

kable(
  table_S7B,
  caption = paste(
    "Table S7B. Observed and model-based FD-positive grapevine",
    "density among the two main grapevine cultivars included in",
    "the micro-scale study (15 m buffer zones along Pinot Noir-",
    "Chasselas boundaries, canton of Vaud). Observed values",
    "correspond to vineyard surface, the total number of FD-positive",
    "grapevines, and the observed FD+ density (FD+ plants ha⁻¹).",
    "Estimated values are expressed as FD-positive grapevines per",
    "hectare with 95% confidence intervals (95% CI). Different letters",
    "indicate significant differences among cultivars."
  )
)
```

```{r}
#| label: table-S7B-export

csv_file_S7B <- "tables/Table_S7B.csv"

write.csv(
  table_S7B,
  csv_file_S7B,
  row.names = FALSE,
  fileEncoding = "UTF-8"
)

cat(
  "CSV successfully created at:\n",
  normalizePath(
    csv_file_S7B,
    winslash = "\\",
    mustWork = TRUE
  )
)
```

