---
title: "Meso-scale flavescence dorée incidence across grapevine cultivars"
author: "Jasmine Cadena"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
editor: source
---

# Analysis overview

This document reproduces the meso-scale analysis of flavescence dorée incidence among the five major grapevine cultivars included in the 500 m buffer zone study in the canton of Vaud.

The response is the number of FD-positive grapevines recorded for each observation. At-risk vineyard surface is included as an exposure offset so that cultivar effects are estimated as FD-positive grapevines per hectare.

# Software and packages

```{r}
#| label: setup

library(glmmTMB)
library(DHARMa)
library(emmeans)
library(multcomp)
library(dplyr)
library(ggplot2)
library(knitr)

dir.create("figures", showWarnings = FALSE)
dir.create("tables", showWarnings = FALSE)
```

# Data import and preparation

## Data import

The public dataset contains the three variables required for the meso-scale analysis: cultivar, at-risk vineyard surface and the number of FD-positive grapevines.

```{r}
#| label: data-import

meso <- read.csv(
  "https://zenodo.org/records/21885421/files/meso.csv?download=1",
  sep = ";",
  stringsAsFactors = FALSE
)

str(meso)
```

## Data preparation

Merlot is set as the reference cultivar.

```{r}
#| label: data-preparation

analysis_data <- meso |>
  filter(
    !is.na(cultivar),
    !is.na(surface_ha),
    !is.na(fd),
    surface_ha > 0
  ) |>
  mutate(
    cultivar = factor(
      cultivar,
      levels = c(
        "Merlot",
        "Pinot Noir",
        "Chasselas",
        "Gamay",
        "Gamaret"
      )
    )
  )

str(analysis_data)
```

# Descriptive statistics

Observed FD-positive grapevine density is calculated from the total number of FD-positive grapevines divided by the total at-risk vineyard surface represented by each cultivar. 

```{r}
#| label: descriptive-statistics

observed_results <- analysis_data |>
  group_by(cultivar) |>
  summarise(
    surface_ha = sum(surface_ha),
    fd_positive = sum(fd),
    observed_density = fd_positive / surface_ha,
    .groups = "drop"
  )

kable(
  observed_results,
  digits = c(0, 0, 0, 2)
)
```

# Family selection

The response variable is the number of FD-positive grapevines recorded for each observation. Observations represent unequal at-risk vineyard surfaces, so `log(surface_ha)` is included as an exposure offset with a fixed coefficient of 1. The expected counts are therefore expressed as FD-positive grapevines per hectare.

A Poisson model is fitted first.

## Poisson model

```{r}
#| label: poisson-model

model_poisson <- glmmTMB(
  fd ~ cultivar + offset(log(surface_ha)),
  data = analysis_data,
  family = poisson(link = "log")
)

summary(model_poisson)
```

Simulation-based residual diagnostics evaluate residual uniformity, dispersion, zero inflation and outliers.

```{r}
#| label: poisson-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the Poisson model."

set.seed(123)

simulation_poisson <- simulateResiduals(
  fittedModel = model_poisson,
  n = 1000
)

plot(simulation_poisson)

testUniformity(simulation_poisson)
testDispersion(simulation_poisson)
testZeroInflation(simulation_poisson)
testOutliers(simulation_poisson)
```

The DHARMa residual diagnostics indicate that the Poisson model does not adequately fit the data. Negative binomial models are therefore evaluated.

## Negative binomial models

NB1 and NB2 negative binomial models differ in how the conditional variance increases with the mean. Both are fitted with the same cultivar effect and exposure offset.

```{r}
#| label: negative-binomial-models

model_nb1 <- glmmTMB(
  fd ~ cultivar + offset(log(surface_ha)),
  data = analysis_data,
  family = nbinom1(link = "log")
)

model_nb2 <- glmmTMB(
  fd ~ cultivar + offset(log(surface_ha)),
  data = analysis_data,
  family = nbinom2(link = "log")
)
```

Simulation-based residual diagnostics are examined separately for each negative binomial model.

### NB1 diagnostics

```{r}
#| label: nb1-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the NB1 negative binomial model."

set.seed(123)

simulation_nb1 <- simulateResiduals(
  fittedModel = model_nb1,
  n = 1000
)

plot(simulation_nb1)

testUniformity(simulation_nb1)
testDispersion(simulation_nb1)
testZeroInflation(simulation_nb1)
testOutliers(simulation_nb1)
```

### NB2 diagnostics

```{r}
#| label: nb2-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the NB2 negative binomial model."

set.seed(123)

simulation_nb2 <- simulateResiduals(
  fittedModel = model_nb2,
  n = 1000
)

plot(simulation_nb2)

testUniformity(simulation_nb2)
testDispersion(simulation_nb2)
testZeroInflation(simulation_nb2)
testOutliers(simulation_nb2)
```

The fitted distributions are compared using Akaike's information criterion. Model selection considers both AIC and simulation-based residual diagnostics.

```{r}
#| label: family-comparison

family_comparison <- data.frame(
  model = c(
    "Poisson",
    "Negative binomial NB1",
    "Negative binomial NB2"
  ),
  AIC = c(
    AIC(model_poisson),
    AIC(model_nb1),
    AIC(model_nb2)
  )
) |>
  arrange(AIC) |>
  mutate(
    delta_AIC = AIC - min(AIC)
  )

kable(
  family_comparison,
  digits = 2
)
```

The NB2 negative binomial model is retained as the final model because it provides satisfactory DHARMa residual diagnostics and the lowest AIC. The zero-inflation test provides no evidence for an excess of zeros, so no zero-inflation component is included in the final model.


# Model structure and term selection

The final model is a negative binomial NB2 model including cultivar as the explanatory variable and at-risk vineyard surface as an exposure offset, allowing FD-positive grapevine density to be estimated per hectare.

# Final model

```{r}
#| label: final-model

final_model <- model_nb2

summary(final_model)
```

# Final-model diagnostics

The NB2 diagnostics presented in the family-selection section assess residual uniformity, dispersion, zero inflation and outliers for the retained model.

# Statistical inference

The overall cultivar effect is evaluated by likelihood-ratio comparison with a model containing the same NB2 family and exposure offset but no cultivar term.

```{r}
#| label: cultivar-likelihood-ratio-test

null_model <- glmmTMB(
  fd ~ offset(log(surface_ha)),
  data = analysis_data,
  family = nbinom2(link = "log")
)

cultivar_lrt <- anova(
  null_model,
  final_model,
  test = "Chisq"
)

cultivar_lrt
```

# Estimated marginal means and contrasts

Estimated marginal means are calculated on the response scale with the offset set to zero. They therefore represent the expected number of FD-positive grapevines per hectare.

```{r}
#| label: estimated-marginal-means

cultivar_emmeans <- emmeans(
  final_model,
  ~ cultivar,
  type = "response",
  offset = 0
)

cultivar_emmeans
```

Pairwise cultivar comparisons use Tukey adjustment for the five-cultivar family of contrasts.

```{r}
#| label: tukey-contrasts

cultivar_contrasts <- pairs(
  cultivar_emmeans,
  adjust = "tukey"
)

cultivar_contrasts
```

Compact-letter groups summarise the Tukey-adjusted comparisons. Cultivars sharing a letter are not significantly different at α = 0.05.

```{r}
#| label: compact-letter-display

cultivar_estimates <- as.data.frame(cultivar_emmeans) |>
  rename(
    estimated_mean = response,
    lower_CL = asymp.LCL,
    upper_CL = asymp.UCL
  )

cultivar_groups <- as.data.frame(
  cld(
    cultivar_emmeans,
    adjust = "tukey",
    Letters = letters
  )
) |>
  transmute(
    cultivar,
    statistical_group = gsub(" ", "", .group)
  )

cultivar_groups
```

# Figures

```{r}
#| label: figure-preparation

cultivar_colors <- c(
  "Chasselas" = "#FCDF77",
  "Pinot Noir" = "#DF8ADB",
  "Merlot" = "#B0C4DE",
  "Gamaret" = "#5B9BD5",
  "Gamay" = "#F8766D"
)

figure_data <- cultivar_estimates |>
  left_join(
    cultivar_groups,
    by = "cultivar"
  ) |>
  arrange(estimated_mean) |>
  mutate(
    cultivar = factor(
      cultivar,
      levels = as.character(cultivar)
    ),
    letter_y = upper_CL + 2
  )

figure_data
```

```{r}
#| label: figure-7
#| fig-cap: "Figure 7 Estimated FD+ plants per hectare for the five main grapevine cultivars included in the 500 m at-risk study zone (Lavaux, Chablais in canton of Vaud). Points represent model-estimated marginal means and error bars indicate 95% confidence intervals. Different letters indicate significant differences among cultivars based on Tukey-adjusted pairwise comparisons (α = 0.05)."
#| fig-width: 7.01
#| fig-height: 4.33

figure_7 <- ggplot(
  figure_data,
  aes(
    x = cultivar,
    y = estimated_mean,
    colour = cultivar
  )
) +
  geom_errorbar(
    aes(
      ymin = lower_CL,
      ymax = upper_CL
    ),
    width = 0.08,
    linewidth = 0.6
  ) +
  geom_point(
    size = 3.4
  ) +
  geom_text(
    aes(
      y = letter_y,
      label = statistical_group
    ),
    colour = "black",
    size = 3.8
  ) +
  scale_colour_manual(
    values = cultivar_colors,
    guide = "none"
  ) +
  scale_y_continuous(
    limits = c(0, 52),
    breaks = seq(0, 50, 10),
    expand = expansion(
      mult = c(0, 0)
    )
  ) +
  labs(
    x = "Cultivar",
    y = "Estimated FD+ plants per ha"
  ) +
  theme_minimal(
    base_size = 10,
    base_family = "Arial"
  ) +
  theme(
    axis.title.x = element_text(
      size = 11,
      colour = "black",
      margin = margin(t = 10)
    ),
    axis.title.y = element_text(
      size = 11,
      colour = "black",
      margin = margin(r = 10)
    ),
    axis.text.x = element_text(
      size = 9,
      colour = "black",
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    axis.text.y = element_text(
      size = 9,
      colour = "black"
    ),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(
      colour = "grey88",
      linewidth = 0.4
    ),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(
      fill = "white",
      colour = NA
    ),
    plot.background = element_rect(
      fill = "white",
      colour = NA
    ),
    plot.margin = margin(
      10,
      12,
      12,
      12
    )
  )

figure_7
```

```{r}
#| label: figure-export

figure_file <- "figures/Figure_7.tiff"

ggsave(
  filename = figure_file,
  plot = figure_7,
  width = 178,
  height = 110,
  units = "mm",
  dpi = 600,
  compression = "lzw",
  bg = "white"
)

cat(
  "Figure files successfully created at:\n",
  normalizePath(
    "figures",
    winslash = "\\",
    mustWork = TRUE
  )
)
```

# Exported tables

## Table S6A

Table S6A reports the parameter estimates of the final NB2 negative binomial model. Merlot is the reference cultivar. Estimates are reported on the log-link scale.

```{r}
#| label: table-S6A

coefficient_matrix <- summary(
  final_model
)$coefficients$cond

parameter_labels <- c(
  "(Intercept)" = "Intercept (Merlot)",
  "cultivarPinot Noir" = "Pinot Noir",
  "cultivarChasselas" = "Chasselas",
  "cultivarGamay" = "Gamay",
  "cultivarGamaret" = "Gamaret"
)

table_S6A <- data.frame(
  Term = unname(
    parameter_labels[
      rownames(coefficient_matrix)
    ]
  ),
  Estimate = round(
    coefficient_matrix[, "Estimate"],
    3
  ),
  SE = round(
    coefficient_matrix[, "Std. Error"],
    3
  ),
  z = round(
    coefficient_matrix[, "z value"],
    3
  ),
  `P value` = ifelse(
    coefficient_matrix[, "Pr(>|z|)"] < 0.001,
    "<0.001",
    sprintf(
      "%.3f",
      coefficient_matrix[, "Pr(>|z|)"]
    )
  ),
  row.names = NULL,
  check.names = FALSE
)

kable(
  table_S6A,
  caption = paste(
    "Table S6A. Parameter estimates of the final NB2 negative",
    "binomial model used to analyse FD-positive grapevine density",
    "in the meso-scale study. Merlot was used as the reference",
    "cultivar. Estimates are reported on the log-link scale."
  )
)
```

```{r}
#| label: table-S6A-export

csv_file_S6A <- "tables/Table_S6A.csv"

write.csv(
  table_S6A,
  csv_file_S6A,
  row.names = FALSE,
  fileEncoding = "UTF-8"
)

cat(
  "CSV successfully created at:\n",
  normalizePath(
    csv_file_S6A,
    winslash = "\\",
    mustWork = TRUE
  )
)
```

## Table S6B

Table S6B combines observed at-risk vineyard surface and FD-positive counts with model-estimated FD-positive grapevine density, 95% confidence intervals and Tukey-adjusted statistical groups.

```{r}
#| label: table-S6B

model_summary <- figure_data |>
  transmute(
    cultivar = as.character(cultivar),
    estimated_mean = estimated_mean,
    lower_CL = lower_CL,
    upper_CL = upper_CL,
    statistical_group = statistical_group
  )

table_S6B <- observed_results |>
  mutate(
    cultivar = as.character(cultivar)
  ) |>
  left_join(
    model_summary,
    by = "cultivar"
  ) |>
  mutate(
    cultivar = factor(
      cultivar,
      levels = c(
        "Chasselas",
        "Merlot",
        "Gamay",
        "Gamaret",
        "Pinot Noir"
      )
    )
  ) |>
  arrange(cultivar) |>
  transmute(
    Cultivar = as.character(cultivar),
    `At-risk surface (ha)` = round(surface_ha, 0),
    `FD+` = round(fd_positive, 0),
    `Observed FD+ ha-1` = round(observed_density, 2),
    `Estimated FD+ ha-1 (95% CI)` = paste0(
      sprintf("%.2f", estimated_mean),
      " (95% CI: ",
      sprintf("%.2f", lower_CL),
      "-",
      sprintf("%.2f", upper_CL),
      ")"
    ),
    `Statistical group` = statistical_group
  )

kable(
  table_S6B,
  caption = paste(
    "Table S6B. Observed and model-based FD-positive grapevine",
    "density among the five main grapevine cultivars included in",
    "the meso-scale study (500 m buffer zones, canton of Vaud).",
    "Observed values correspond to at-risk vineyard surface (ha),",
    "the total number of FD-positive grapevines, and observed FD+",
    "density. Estimated values are expressed per hectare with 95%",
    "confidence intervals. Different letters indicate significant",
    "differences based on Tukey-adjusted pairwise comparisons",
    "(α = 0.05)."
  ),
  align = c("l", "r", "r", "r", "l", "c"),
  format.args = list(big.mark = ",")
)
```

```{r}
#| label: table-S6B-export

csv_file_S6B <- "tables/Table_S6B.csv"

write.csv(
  table_S6B,
  csv_file_S6B,
  row.names = FALSE,
  fileEncoding = "UTF-8"
)

cat(
  "CSV successfully created at:\n",
  normalizePath(
    csv_file_S6B,
    winslash = "\\",
    mustWork = TRUE
  )
)
```

