---
title: "Distribution of FD phytoplasma infection within grapevine plants"
author: "Jasmine Cadena"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
editor: source
---

# Analysis overview

This document reproduces the descriptive analysis of FD phytoplasma detection in different leaf positions of infected grapevine plants following controlled transmission experiments. The analysis summarises the proportion of sampled leaf positions that tested positive within each cultivar.



# Software and packages

```{r}
#| label: setup

if (getRversion() < "4.5.2") {
  warning(
    "This analysis was developed using R 4.5.2. Results should be checked when using an older R version."
  )
}

required_packages <- c(
  "dplyr",
  "ggplot2",
  "scales",
  "knitr"
)

missing_packages <- required_packages[
  !vapply(
    required_packages,
    requireNamespace,
    logical(1),
    quietly = TRUE
  )
]

if (length(missing_packages) > 0) {
  stop(
    paste0(
      "The following R packages are required but not installed: ",
      paste(missing_packages, collapse = ", "),
      ". Install them with install.packages() before rendering this document."
    )
  )
}

library(dplyr)
library(ggplot2)
library(scales)
library(knitr)

dir.create("figures", showWarnings = FALSE)
dir.create("tables", showWarnings = FALSE)
```

# Data import and preparation

## Data import

The public dataset contains one row per recorded leaf position. 

```{r}
#| label: data-import

plant_distribution <- read.csv2(
  "https://zenodo.org/records/21885421/files/plant_distribution.csv?download=1",
  stringsAsFactors = FALSE
)

str(plant_distribution)
```


## Data preparation

The public dataset is already restricted to inoculated plants that tested positive at the whole-plant level. The preparation step standardises factor order and converts cultivar codes into the names displayed in the figure.

```{r}
#| label: data-preparation

analysis_data <- plant_distribution |>
  mutate(
    cultivar = factor(
      cultivar,
      levels = c("ME", "PN", "CH", "GY", "GM", "CR", "CS")
    ),
    cultivar_name = recode(
      cultivar,
      "CH" = "Chasselas",
      "PN" = "Pinot Noir",
      "ME" = "Merlot",
      "CR" = "Chardonnay",
      "GM" = "Gamaret",
      "GY" = "Gamay",
      "CS" = "Cabernet Sauvignon"
    ),
    cultivar_name = factor(
      cultivar_name,
      levels = c(
        "Merlot",
        "Pinot Noir",
        "Chasselas",
        "Gamay",
        "Gamaret",
        "Chardonnay",
        "Cabernet Sauvignon"
      )
    ),
    part = factor(
      part,
      levels = c("BL", "IL", "PUL", "DUL"),
      labels = c(
        "Basal leaf",
        "Inoculated leaf",
        "Proximal upper leaf",
        "Distal upper leaf"
      )
    )
  )

str(analysis_data)
```

# Descriptive statistics

The percentage of infected observations was calculated as the number of infected observations divided by the total number of observations available for each cultivar × leaf-part combination.

```{r}
#| label: descriptive-summary

plant_distribution_summary <- analysis_data |>
  group_by(cultivar, cultivar_name, part) |>
  summarise(
    infected_leaf_positions = sum(leaf_status == "infected"),
    sampled_leaf_positions = n(),
    infected_proportion =
      infected_leaf_positions / sampled_leaf_positions,
    .groups = "drop"
  ) |>
  mutate(
    infected_sampled = paste0(
      infected_leaf_positions,
      "/",
      sampled_leaf_positions
    )
  )

knitr::kable(
  plant_distribution_summary |>
    dplyr::select(
      cultivar_name,
      part,
      infected_leaf_positions,
      sampled_leaf_positions,
      infected_proportion
    ),
  digits = 3,
  col.names = c(
    "Cultivar",
    "Leaf position",
    "Infected",
    "Sampled",
    "Proportion infected"
  )
)
```

# Figure

Bars show the observed proportion of infected leaf positions among all sampled leaf positions for each cultivar. Labels above bars report the number of infected observations over the total number sampled.

```{r}
#| label: plant-distribution-figure
#| fig-cap: "Distribution of FD phytoplasma detection among leaf positions of infected grapevine plants. Bars represent the proportion of sampled leaf positions that tested positive for FD phytoplasma. Numbers above bars indicate the number of infected observations over the total number sampled."
#| fig-width: 7.01
#| fig-height: 5.91

cultivar_colours <- c(
  "Merlot" = "#B0C4DE",
  "Pinot Noir" = "#DF8ADB",
  "Chasselas" = "#E4D00A",
  "Gamay" = "#F8766D",
  "Gamaret" = "#5B9BD5",
  "Chardonnay" = "#A6D854",
  "Cabernet Sauvignon" = "#F4A32C"
)

plant_distribution_figure <- ggplot(
  plant_distribution_summary,
  aes(
    x = cultivar_name,
    y = infected_proportion,
    fill = cultivar_name
  )
) +
  geom_col(
    colour = "black",
    alpha = 0.8,
    width = 0.7
  ) +
  geom_text(
    aes(label = infected_sampled),
    vjust = -0.35,
    size = 3.8
  ) +
  facet_wrap(
    ~ part,
    ncol = 2
  ) +
  scale_fill_manual(
    values = cultivar_colours,
    drop = FALSE
  ) +
  scale_y_continuous(
    limits = c(0, 1.08),
    breaks = seq(0, 1, by = 0.2),
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    x = "Cultivar",
    y = "Proportion of infected leaves"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    text = element_text(family = "Arial"),
    panel.border = element_rect(
      colour = "grey30",
      fill = NA,
      linewidth = 0.8
    ),
    strip.background = element_rect(
      fill = "white",
      colour = "grey30",
      linewidth = 0.8
    ),
    axis.text.x = element_text(
      size = 9,
      angle = 45,
      hjust = 1,
      colour = "black"
    ),
    axis.text.y = element_text(
      size = 9,
      colour = "black"
    ),
    axis.title.x = element_text(
      size = 11,
      colour = "black"
    ),
    axis.title.y = element_text(
      size = 11,
      colour = "black"
    ),
    strip.text = element_text(
      size = 11,
      face = "bold"
    ),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(
      colour = "grey85",
      linewidth = 0.3
    ),
    panel.background = element_rect(
      fill = "white",
      colour = NA
    ),
    plot.background = element_rect(
      fill = "white",
      colour = NA
    )
  )

plant_distribution_figure

tiff_file <- "figures/Figure 4.tif"

ggsave(
  filename = tiff_file,
  plot = plant_distribution_figure,
  device = "tiff",
  width = 178,
  height = 150,
  units = "mm",
  dpi = 600,
  compression = "lzw",
  bg = "white"
)

cat(
  "TIFF successfully created at:\n",
  normalizePath(
    tiff_file,
    winslash = "\\",
    mustWork = TRUE
  )
)
```

