---
title: "Relative FD phytoplasma titre in field samples"
author: "Jasmine Cadena"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
editor: source
---

# Analysis overview

This analysis examines relative Flavescence dorée phytoplasma (FDp) titre in field samples of Merlot, Chasselas and Pinot Noir. Relative titre is expressed as ΔΔCt-derived fold change, with Merlot as the reference cultivar.

# Software and packages

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
library(glmmTMB)
library(DHARMa)
library(emmeans)
library(multcomp)
library(multcompView)
library(knitr)

dir.create("figures", showWarnings = FALSE)
dir.create("tables", showWarnings = FALSE)
```

# Data import and preparation

## Data import

The dataset contains the cultivar and the ΔΔCt-derived fold change for each field sample.

```{r}
#| label: data-import

field_titre <- read.csv2(
  "https://zenodo.org/records/21885421/files/field_titre.csv?download=1",
  dec = ".",
  stringsAsFactors = FALSE
)

str(field_titre)
```

## Data preparation

The dataset is imported and cultivar is encoded as a factor before fitting the candidate models.

```{r}
#| label: data-preparation

analysis_data <- field_titre |>
  mutate(
    cultivar = factor(
      cultivar,
      levels = c(
        "Merlot",
        "Chasselas",
        "Pinot Noir"
      )
    )
  )

summary(analysis_data$fold_change)
table(analysis_data$cultivar)
```

# Descriptive statistics

Descriptive statistics summarise the observed fold-change distribution for each cultivar.

```{r}
#| label: descriptive-statistics

observed_summary <- analysis_data |>
  group_by(cultivar) |>
  summarise(
    n = n(),
    mean = mean(fold_change),
    median = median(fold_change),
    sd = sd(fold_change),
    minimum = min(fold_change),
    maximum = max(fold_change),
    .groups = "drop"
  )

kable(
  observed_summary,
  digits = 2
)
```

# Family selection

Fold-change values are positive and right-skewed. Gaussian models on the original and logarithmic scales and a Gamma model with a log link are therefore compared. All three models contain cultivar as the explanatory variable.

## Gaussian model on the original scale

```{r}
#| label: gaussian-model

model_gaussian <- glmmTMB(
  fold_change ~ cultivar,
  family = gaussian(link = "identity"),
  data = analysis_data
)

summary(model_gaussian)
```

```{r}
#| label: gaussian-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the Gaussian model fitted on the original response scale."

set.seed(1234)

simulation_gaussian <- simulateResiduals(
  fittedModel = model_gaussian,
  n = 1000
)

plot(simulation_gaussian)

gaussian_uniformity <- testUniformity(simulation_gaussian)
gaussian_dispersion <- testDispersion(simulation_gaussian)
gaussian_outliers <- testOutliers(simulation_gaussian)

gaussian_uniformity
gaussian_dispersion
gaussian_outliers

plotResiduals(
  simulation_gaussian,
  form = analysis_data$cultivar
)
```

## Log-Gaussian model

```{r}
#| label: log-gaussian-model

model_log_gaussian <- glmmTMB(
  log(fold_change) ~ cultivar,
  family = gaussian(link = "identity"),
  data = analysis_data
)

summary(model_log_gaussian)
```

```{r}
#| label: log-gaussian-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the Gaussian model fitted on log-transformed fold-change values."

set.seed(1234)

simulation_log_gaussian <- simulateResiduals(
  fittedModel = model_log_gaussian,
  n = 1000
)

plot(simulation_log_gaussian)

log_gaussian_uniformity <- testUniformity(simulation_log_gaussian)
log_gaussian_dispersion <- testDispersion(simulation_log_gaussian)
log_gaussian_outliers <- testOutliers(simulation_log_gaussian)

log_gaussian_uniformity
log_gaussian_dispersion
log_gaussian_outliers

plotResiduals(
  simulation_log_gaussian,
  form = analysis_data$cultivar
)
```

## Gamma model with a log link

```{r}
#| label: gamma-model

model_gamma <- glmmTMB(
  fold_change ~ cultivar,
  family = Gamma(link = "log"),
  data = analysis_data
)

summary(model_gamma)
```

```{r}
#| label: gamma-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the Gamma model with a log link."

set.seed(1234)

simulation_gamma <- simulateResiduals(
  fittedModel = model_gamma,
  n = 1000
)

plot(simulation_gamma)

gamma_uniformity <- testUniformity(simulation_gamma)
gamma_dispersion <- testDispersion(simulation_gamma)
gamma_outliers <- testOutliers(simulation_gamma)

gamma_uniformity
gamma_dispersion
gamma_outliers

plotResiduals(
  simulation_gamma,
  form = analysis_data$cultivar
)
```

The log-Gaussian model is fitted to a transformed response, whereas the Gaussian and Gamma models are fitted directly to fold change. The Jacobian correction places the log-Gaussian AIC on the original response scale before the models are compared.

```{r}
#| label: family-comparison

AIC_gaussian <- AIC(model_gaussian)

AIC_log_gaussian_corrected <-
  AIC(model_log_gaussian) +
  2 * sum(log(analysis_data$fold_change))

AIC_gamma <- AIC(model_gamma)

family_comparison <- data.frame(
  distribution = c(
    "Gaussian",
    "Log-Gaussian",
    "Gamma with log link"
  ),
  AIC = c(
    AIC_gaussian,
    AIC_log_gaussian_corrected,
    AIC_gamma
  ),
  uniformity_p = c(
    gaussian_uniformity$p.value,
    log_gaussian_uniformity$p.value,
    gamma_uniformity$p.value
  ),
  dispersion_p = c(
    gaussian_dispersion$p.value,
    log_gaussian_dispersion$p.value,
    gamma_dispersion$p.value
  ),
  outliers_p = c(
    gaussian_outliers$p.value,
    log_gaussian_outliers$p.value,
    gamma_outliers$p.value
  )
) |>
  arrange(AIC) |>
  mutate(
    delta_AIC = AIC - min(AIC)
  )

kable(
  family_comparison,
  digits = 3
)
```

The log-Gaussian model is retained because it provides the lowest AIC and satisfactory residual diagnostics. Although the DHARMa outlier test is significant (P = 0.049), only two observations are identified as potential outliers and are retained because they correspond to valid observations.

# Model structure and term selection

The final model includes cultivar as the explanatory variable in a Gaussian model fitted to log-transformed fold-change values.

# Final model

```{r}
#| label: final-model

final_model <- model_log_gaussian

summary(final_model)
```

# Final-model diagnostics

The log-Gaussian diagnostics presented in the family-selection section assess residual uniformity, dispersion and outliers for the retained model.

# Statistical inference

The overall cultivar effect is evaluated by likelihood-ratio comparison with a log-Gaussian model containing no cultivar term.

```{r}
#| label: cultivar-likelihood-ratio-test

null_model <- glmmTMB(
  log(fold_change) ~ 1,
  family = gaussian(link = "identity"),
  data = analysis_data
)

cultivar_lrt <- anova(
  null_model,
  final_model,
  test = "Chisq"
)

cultivar_lrt
```

# Estimated marginal means and contrasts

Estimated marginal means are calculated for each cultivar and back-transformed to the fold-change scale.

```{r}
#| label: estimated-marginal-means

cultivar_emmeans <- emmeans(
  final_model,
  ~ cultivar
)

cultivar_emmeans_response <- summary(
  cultivar_emmeans,
  type = "response",
  infer = c(TRUE, TRUE)
) |>
  as.data.frame()

cultivar_emmeans_response
```

Pairwise cultivar comparisons use Tukey adjustment for the three-cultivar family of contrasts.

```{r}
#| label: tukey-contrasts

cultivar_contrasts <- contrast(
  cultivar_emmeans,
  method = "pairwise",
  adjust = "tukey"
)

cultivar_contrasts_response <- summary(
  cultivar_contrasts,
  type = "response",
  infer = c(TRUE, TRUE)
)

cultivar_contrasts_response
```

Compact-letter groups summarise the Tukey-adjusted comparisons. Cultivars sharing a letter are not significantly different at α = 0.05.

```{r}
#| label: compact-letter-display

cultivar_groups <- cld(
  cultivar_emmeans,
  adjust = "tukey",
  Letters = letters,
  type = "response"
) |>
  as.data.frame() |>
  mutate(
    .group = trimws(.group)
  )

cultivar_groups
```

# Figures

The figure uses the back-transformed estimated marginal means, 95% confidence intervals, Tukey groups and observed sample sizes.

```{r}
#| label: figure-preparation

figure_data <- cultivar_emmeans_response |>
  dplyr::select(
    cultivar,
    response,
    lower.CL,
    upper.CL
  ) |>
  left_join(
    cultivar_groups |>
      dplyr::select(
        cultivar,
        .group
      ),
    by = "cultivar"
  ) |>
  transmute(
    cultivar,
    estimate = response,
    lower = lower.CL,
    upper = upper.CL,
    letters = .group
  ) |>
  left_join(
    observed_summary |>
      dplyr::select(
        cultivar,
        n
      ),
    by = "cultivar"
  ) |>
  mutate(
    cultivar = factor(
      cultivar,
      levels = c(
        "Merlot",
        "Chasselas",
        "Pinot Noir"
      )
    ),
    n_label = paste0(
      "n = ",
      n
    ),
    letter_y = upper * 1.20,
    n_y = 0.40
  )

figure_data
```

```{r}
#| label: figure-10
#| fig-cap: "Figure 10 Model-estimated FDp titre expressed as fold change relative to Merlot in field samples of Chasselas and Pinot Noir. Points represent estimated marginal means obtained from a Gaussian model fitted on log-transformed data, and error bars indicate 95% confidence intervals. Fold changes were calculated using the ΔΔCt method, with Merlot used as the reference cultivar. Different letters indicate significant differences among cultivars based on Tukey-adjusted pairwise comparisons (α = 0.05)."
#| fig-width: 7.01
#| fig-height: 4.33

cultivar_colours <- c(
  "Merlot" = "#B0C4DE",
  "Chasselas" = "#E4D00A",
  "Pinot Noir" = "#DF8ADB"
)

cultivar_labels <- c(
  "Merlot" = "Merlot",
  "Chasselas" = "Chasselas",
  "Pinot Noir" = "Pinot Noir"
)

figure_10 <- ggplot(
  figure_data,
  aes(
    x = cultivar,
    y = estimate,
    colour = cultivar
  )
) +
  geom_errorbar(
    aes(
      ymin = lower,
      ymax = upper
    ),
    width = 0.08,
    linewidth = 0.6
  ) +
  geom_point(
    size = 3.4
  ) +
  geom_text(
    aes(
      y = letter_y,
      label = letters
    ),
    colour = "black",
    size = 4.2,
    fontface = "plain"
  ) +
  geom_text(
    aes(
      y = n_y,
      label = n_label
    ),
    colour = "black",
    size = 3.2,
    fontface = "plain",
    vjust = 0
  ) +
  scale_colour_manual(
    values = cultivar_colours,
    guide = "none"
  ) +
  scale_x_discrete(
    labels = cultivar_labels
  ) +
  scale_y_log10(
    limits = c(
      0.35,
      max(figure_data$letter_y) * 1.15
    ),
    breaks = c(
      0.5,
      1,
      2,
      5,
      10,
      20,
      50
    ),
    labels = c(
      "0.5",
      "1",
      "2",
      "5",
      "10",
      "20",
      "50"
    ),
    expand = expansion(
      mult = c(
        0,
        0.02
      )
    )
  ) +
  labs(
    x = "Cultivar",
    y = "Estimated relative FD phytoplasma titre"
  ) +
  theme_minimal(
    base_size = 10,
    base_family = "Arial"
  ) +
  theme(
    axis.title.x = element_text(
      size = 11,
      colour = "black",
      margin = margin(
        t = 10
      )
    ),
    axis.title.y = element_text(
      size = 11,
      colour = "black",
      margin = margin(
        r = 10
      )
    ),
    axis.text.x = element_text(
      size = 9,
      colour = "black",
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    axis.text.y = element_text(
      size = 9,
      colour = "black"
    ),
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(
      colour = "grey88",
      linewidth = 0.4
    ),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_blank(),
    panel.background = element_rect(
      fill = "white",
      colour = NA
    ),
    plot.background = element_rect(
      fill = "white",
      colour = NA
    ),
    plot.margin = margin(
      10,
      12,
      12,
      12
    )
  )

figure_10
```

```{r}
#| label: figure-export

figure_file <- "figures/Figure_10.tiff"

# The figure is exported with the same dimensions and graphical settings used in the original analysis.
ggsave(
  filename = figure_file,
  plot = figure_10,
  device = "tiff",
  width = 178,
  height = 110,
  units = "mm",
  dpi = 600,
  compression = "lzw",
  bg = "white"
)

cat(
  "TIFF successfully created at:\n",
  normalizePath(
    figure_file,
    winslash = "\\",
    mustWork = TRUE
  )
)
```

# Exported tables

**Table Supplementary 9. Model-based relative FD phytoplasma titre in field samples of Merlot, Chasselas and Pinot Noir. Relative FD phytoplasma titre is expressed as fold change relative to Merlot using the ΔΔCt method.**

## Table S8A

Table S8A reports the parameter estimates of the final log-Gaussian model. Merlot is the reference cultivar. Estimates are reported on the log-transformed response scale.

```{r}
#| label: table-S8A

coefficient_matrix <- summary(
  final_model
)$coefficients$cond

parameter_labels <- c(
  "(Intercept)" = "Intercept (Merlot)",
  "cultivarChasselas" = "Chasselas",
  "cultivarPinot Noir" = "Pinot Noir"
)

table_S8A <- data.frame(
  Term = unname(
    parameter_labels[
      rownames(coefficient_matrix)
    ]
  ),
  Estimate = sprintf(
    "%.3f",
    coefficient_matrix[, "Estimate"]
  ),
  SE = sprintf(
    "%.3f",
    coefficient_matrix[, "Std. Error"]
  ),
  z = sprintf(
    "%.3f",
    coefficient_matrix[, "z value"]
  ),
  `P value` = ifelse(
    coefficient_matrix[, "Pr(>|z|)"] < 0.001,
    "<0.001",
    sprintf(
      "%.3f",
      coefficient_matrix[, "Pr(>|z|)"]
    )
  ),
  check.names = FALSE,
  row.names = NULL
)

kable(
  table_S8A,
  align = c("l", "r", "r", "r", "r"),
  caption = paste(
    "Table S8A. Parameter estimates of the final log-Gaussian",
    "model. Merlot is the reference cultivar. Estimates are",
    "reported on the log-transformed response scale."
  )
)
```

```{r}
#| label: table-S8A-export

csv_file_S8A <- "tables/Table_S8A.csv"

write.csv(
  table_S8A,
  csv_file_S8A,
  row.names = FALSE,
  fileEncoding = "UTF-8",
  quote = FALSE
)

cat(
  "CSV successfully created at:\n",
  normalizePath(
    csv_file_S8A,
    winslash = "\\",
    mustWork = TRUE
  )
)
```

## Table S8B

Table S8B reports estimated marginal means back-transformed to the response scale and expressed as ΔΔCt-derived fold change relative to Merlot with 95% confidence intervals. Different letters indicate significant differences among cultivars based on Tukey-adjusted pairwise comparisons.

```{r}
#| label: table-S8B

table_S8B <- cultivar_emmeans_response |>
  dplyr::select(
    cultivar,
    response,
    lower.CL,
    upper.CL
  ) |>
  left_join(
    cultivar_groups |>
      dplyr::select(
        cultivar,
        .group
      ),
    by = "cultivar"
  ) |>
  transmute(
    Cultivar = case_when(
      cultivar == "Merlot" ~ "Merlot (reference)",
      TRUE ~ as.character(cultivar)
    ),
    `Estimated fold change (95% CI)` = sprintf(
      "%.2f (%.2f-%.2f)",
      response,
      lower.CL,
      upper.CL
    ),
    `Statistical group` = .group
  )

kable(
  table_S8B,
  align = c("l", "c", "c"),
  caption = paste(
    "Table S8B. Estimated marginal means back-transformed to the",
    "response scale and expressed as ΔΔCt-derived fold change",
    "relative to Merlot with 95% confidence intervals (95% CI).",
    "Different letters indicate significant differences among",
    "cultivars based on Tukey-adjusted pairwise comparisons."
  )
)
```

```{r}
#| label: table-S8B-export

csv_file_S8B <- "tables/Table_S8B.csv"

write.csv(
  table_S8B,
  csv_file_S8B,
  row.names = FALSE,
  fileEncoding = "UTF-8",
  quote = FALSE
)

cat(
  "CSV successfully created at:\n",
  normalizePath(
    csv_file_S8B,
    winslash = "\\",
    mustWork = TRUE
  )
)
```

