---
title: "Macro-scale flavescence dorée incidence across grapevine cultivars"
author: "Jasmine Cadena"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
editor: source
---

# Analysis overview

This document reproduces the macro-scale analysis of flavescence dorée incidence across the five major grapevine cultivars included in the final analysis of Lavaux, Chablais and Central Valais.

The response is the number of FD-positive grapevines recorded for each vineyard unit. Vineyard surface is included as an exposure offset so that cultivar effects are estimated as FD-positive grapevines per hectare.

# Software and packages

```{r}
#| label: setup

library(glmmTMB)
library(DHARMa)
library(emmeans)
library(multcomp)
library(dplyr)
library(ggplot2)
library(knitr)

dir.create("figures", showWarnings = FALSE)
dir.create("tables", showWarnings = FALSE)
```

# Data import and preparation

## Data import

The public dataset contains only the five major grapevine cultivars included in the final macro-scale analysis: Chasselas, Pinot Noir, Gamay, Sylvaner and Arvine.

```{r}
#| label: data-import

macro <- read.csv2(
  "https://zenodo.org/records/21885421/files/macro.csv?download=1",
  dec = ".",
  stringsAsFactors = FALSE
)

str(macro)
```

## Data preparation

Chasselas is set as the reference cultivar.

```{r}
#| label: data-preparation

analysis_data <- macro |>
  filter(
    !is.na(cultivar),
    !is.na(surface_ha),
    !is.na(fd),
    surface_ha > 0
  ) |>
  mutate(
    cultivar = factor(
      cultivar,
      levels = c(
        "Chasselas",
        "Pinot Noir",
        "Gamay",
        "Sylvaner",
        "Arvine"
      )
    )
  )

str(analysis_data)
```

# Descriptive statistics

Observed FD-positive grapevine density is calculated from the total number of FD-positive grapevines divided by the total vineyard surface represented by each cultivar.

```{r}
#| label: descriptive-statistics

observed_summary <- analysis_data |>
  group_by(cultivar) |>
  summarise(
    surface_ha = sum(surface_ha),
    fd_positive = sum(fd),
    observed_fd_ha = fd_positive / surface_ha,
    .groups = "drop"
  )

kable(
  observed_summary,
  digits = c(0, 0, 0, 2)
)
```

# Family selection

The response variable is the number of FD-positive grapevines recorded in each vineyard unit. Vineyard units differ in size, so `log(surface_ha)` is included as an offset with a fixed coefficient of 1. This expresses the expected number of FD-positive grapevines per hectare.

A Poisson model is fitted first.

## Poisson model

```{r}
#| label: poisson-model

model_poisson <- glmmTMB(
  fd ~ cultivar + offset(log(surface_ha)),
  data = analysis_data,
  family = poisson(link = "log")
)
```

```{r}
#| label: poisson-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the Poisson model."

set.seed(123)

simulation_poisson <- simulateResiduals(
  fittedModel = model_poisson,
  n = 1000
)

plot(simulation_poisson)

testUniformity(simulation_poisson)
testDispersion(simulation_poisson)
testZeroInflation(simulation_poisson)
testOutliers(simulation_poisson)
```

Model diagnostics indicate that the Poisson distribution does not adequately fit the data. Negative binomial models are therefore evaluated.

## Negative binomial and zero-inflated models

NB1 and NB2 negative binomial models differ in how the conditional variance increases with the mean. Both are fitted with the same cultivar effect and exposure offset. 

```{r}
#| label: alternative-count-models

model_nb1 <- glmmTMB(
  fd ~ cultivar + offset(log(surface_ha)),
  data = analysis_data,
  family = nbinom1(link = "log")
)

model_nb2 <- glmmTMB(
  fd ~ cultivar + offset(log(surface_ha)),
  data = analysis_data,
  family = nbinom2(link = "log")
)

```

Simulation-based diagnostics evaluate residual uniformity, dispersion, zero inflation and outliers for each model.

### NB1 diagnostics

```{r}
#| label: nb1-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the NB1 negative binomial model."

set.seed(123)

simulation_nb1 <- simulateResiduals(
  fittedModel = model_nb1,
  n = 1000
)

plot(simulation_nb1)

testUniformity(simulation_nb1)
testDispersion(simulation_nb1)
testZeroInflation(simulation_nb1)
testOutliers(simulation_nb1)
```

### NB2 diagnostics

```{r}
#| label: nb2-diagnostics
#| fig-cap: "DHARMa residual diagnostics for the NB2 negative binomial model."

set.seed(123)

simulation_nb2 <- simulateResiduals(
  fittedModel = model_nb2,
  n = 1000
)

plot(simulation_nb2)

testUniformity(simulation_nb2)
testDispersion(simulation_nb2)
testZeroInflation(simulation_nb2)
testOutliers(simulation_nb2)
```


The fitted distributions are compared using Akaike's information criterion. Model selection considers both AIC and simulation-based residual diagnostics.

```{r}
#| label: family-comparison

family_comparison <- data.frame(
  model = c(
    "Poisson",
    "Negative binomial NB1",
    "Negative binomial NB2"
  ),
  AIC = c(
    AIC(model_poisson),
    AIC(model_nb1),
    AIC(model_nb2)
  )
) |>
  arrange(AIC) |>
  mutate(
    delta_AIC = AIC - min(AIC)
  )

kable(
  family_comparison,
  digits = 2
)
```
The NB2 model is retained as the final model because it provides satisfactory simulation-based residual diagnostics. Although the NB1 model has a lower AIC, it shows significant residual dispersion, whereas the NB2 model adequately captures the mean-variance relationship. The zero-inflation test provides no evidence for an excess of zeros, so no zero-inflation component is included in the final model.


# Model structure and term selection

The final model is a negative binomial NB2 model including cultivar as the explanatory variable and at-risk vineyard surface as an exposure offset, allowing FD-positive grapevine density to be estimated per hectare.

# Final model

```{r}
#| label: final-model

final_model <- model_nb2

summary(final_model)
```

# Final-model diagnostics

The NB2 diagnostics presented in the family-selection section assess residual uniformity, dispersion, zero inflation and outliers for the retained model.

# Statistical inference

The overall cultivar effect is evaluated by likelihood-ratio comparison with a model containing the same NB2 family and exposure offset, but no cultivar term.

```{r}
#| label: cultivar-likelihood-ratio-test

null_model <- glmmTMB(
  fd ~ offset(log(surface_ha)),
  data = analysis_data,
  family = nbinom2(link = "log")
)

cultivar_lrt <- anova(
  null_model,
  final_model,
  test = "Chisq"
)

cultivar_lrt
```

# Estimated marginal means and contrasts

Estimated marginal means are calculated on the response scale with the offset set to zero. They therefore represent the expected number of FD-positive grapevines per hectare.

```{r}
#| label: estimated-marginal-means

cultivar_emmeans <- emmeans(
  final_model,
  ~ cultivar,
  type = "response",
  offset = 0
)

cultivar_emmeans
```

Pairwise cultivar comparisons use Tukey adjustment for the five-cultivar family of contrasts.

```{r}
#| label: tukey-contrasts

cultivar_contrasts <- pairs(
  cultivar_emmeans,
  adjust = "tukey"
)

cultivar_contrasts
```

Compact-letter groups summarise the Tukey-adjusted comparisons. Cultivars sharing a letter are not significantly different at α = 0.05.

```{r}
#| label: compact-letter-display

cultivar_estimates <- as.data.frame(cultivar_emmeans) |>
  rename(
    estimated_mean = response,
    lower_CL = asymp.LCL,
    upper_CL = asymp.UCL
  )

cultivar_groups <- as.data.frame(
  cld(
    cultivar_emmeans,
    adjust = "tukey",
    Letters = letters
  )
) |>
  transmute(
    cultivar,
    statistical_group = gsub(" ", "", .group)
  )

cultivar_groups
```

# Figures

```{r}
#| label: figure-preparation

cultivar_colors <- c(
  "Chasselas" = "#FCDF77",
  "Pinot Noir" = "#DF8ADB",
  "Gamay" = "#F8766D",
  "Sylvaner" = "#7CAE00",
  "Arvine" = "#00BFC4"
)

figure_data <- cultivar_estimates |>
  left_join(
    cultivar_groups,
    by = "cultivar"
  ) |>
  arrange(estimated_mean) |>
  mutate(
    cultivar = factor(
      cultivar,
      levels = as.character(cultivar)
    ),
    letter_y = upper_CL * 1.25
  )

figure_data
```

```{r}
#| label: figure-6
#| fig-cap: "Figure 6 Estimated FD+ plants per hectare for the five main grapevine cultivars included in the macro study zone (Lavaux, Chablais and Central Valais). Points represent model-estimated marginal means and error bars indicate 95% confidence intervals. Different letters indicate significant differences among cultivars based on Tukey-adjusted pairwise comparisons (α = 0.05)."
#| fig-width: 7.01
#| fig-height: 4.33

figure_6 <- ggplot(
  figure_data,
  aes(
    x = cultivar,
    y = estimated_mean,
    colour = cultivar
  )
) +
  geom_errorbar(
    aes(
      ymin = lower_CL,
      ymax = upper_CL
    ),
    width = 0.08,
    linewidth = 0.6
  ) +
  geom_point(
    size = 3.4
  ) +
  geom_text(
    aes(
      y = letter_y,
      label = statistical_group
    ),
    colour = "black",
    size = 3.8
  ) +
  scale_colour_manual(
    values = cultivar_colors,
    guide = "none"
  ) +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(
      base = 10,
      sigma = 0.15
    ),
    breaks = c(
      0,
      0.05,
      0.1,
      0.2,
      0.5,
      1,
      2,
      5,
      8,
      10
    ),
    limits = c(-0.06, 12),
    expand = expansion(
      mult = c(0.04, 0.10)
    )
  ) +
  labs(
    x = "Cultivar",
    y = "Estimated FD+ plants per ha"
  ) +
  theme_minimal(
    base_size = 10,
    base_family = "Arial"
  ) +
  theme(
    axis.title.x = element_text(
      size = 11,
      colour = "black",
      margin = margin(t = 10)
    ),
    axis.title.y = element_text(
      size = 11,
      colour = "black",
      margin = margin(r = 10)
    ),
    axis.text.x = element_text(
      size = 9,
      colour = "black",
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    axis.text.y = element_text(
      size = 9,
      colour = "black"
    ),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(
      colour = "grey88",
      linewidth = 0.4
    ),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(
      fill = "white",
      colour = NA
    ),
    plot.background = element_rect(
      fill = "white",
      colour = NA
    ),
    plot.margin = margin(
      10,
      12,
      12,
      12
    )
  )

figure_6
```

```{r}
#| label: figure-export

figure_file <- "figures/Figure_6.tiff"

ggsave(
  filename = figure_file,
  plot = figure_6,
  width = 178,
  height = 110,
  units = "mm",
  dpi = 600,
  compression = "lzw",
  bg = "white"
)

cat(
  "TIFF successfully created at:\n",
  normalizePath(
    figure_file,
    winslash = "\\",
    mustWork = TRUE
  )
)
```

# Exported tables

## Table S5A

Table S5A reports the parameter estimates of the final NB2 negative binomial model. Chasselas is the reference cultivar. Estimates are reported on the log-link scale.

```{r}
#| label: table-s5a

coefficient_matrix <- summary(
  final_model
)$coefficients$cond

parameter_labels <- c(
  "(Intercept)" = "Intercept (Chasselas)",
  "cultivarPinot Noir" = "Pinot Noir",
  "cultivarGamay" = "Gamay",
  "cultivarSylvaner" = "Sylvaner",
  "cultivarArvine" = "Arvine"
)

table_s5a <- data.frame(
  Parameter = unname(
    parameter_labels[
      rownames(coefficient_matrix)
    ]
  ),
  Estimate = round(
    coefficient_matrix[, "Estimate"],
    3
  ),
  SE = round(
    coefficient_matrix[, "Std. Error"],
    3
  ),
  z = round(
    coefficient_matrix[, "z value"],
    3
  ),
  P = ifelse(
    coefficient_matrix[, "Pr(>|z|)"] < 0.001,
    "<0.001",
    sprintf(
      "%.3f",
      coefficient_matrix[, "Pr(>|z|)"]
    )
  ),
  row.names = NULL
)

kable(
  table_s5a,
  caption = paste(
    "Table S5A. Parameter estimates of the final NB2 negative",
    "binomial model used to analyse the density of FD-positive",
    "grapevines in the macro-scale analysis. Chasselas is the",
    "reference cultivar. Estimates are reported on the log-link",
    "scale."
  )
)
```

```{r}
#| label: table-s5a-export

csv_file_s5a <- "tables/Table_S5A.csv"

write.csv(
  table_s5a,
  csv_file_s5a,
  row.names = FALSE,
  fileEncoding = "UTF-8"
)

cat(
  "CSV successfully created at:\n",
  normalizePath(
    csv_file_s5a,
    winslash = "\\",
    mustWork = TRUE
  )
)
```

## Table S5B

Table S5B combines observed vineyard surface and FD-positive counts with model-estimated FD-positive grapevine density, 95% confidence intervals and Tukey-adjusted statistical groups.

```{r}
#| label: table-S5B

model_summary <- figure_data |>
  transmute(
    cultivar = as.character(cultivar),
    estimated_fd_ha = estimated_mean,
    lower_CL = lower_CL,
    upper_CL = upper_CL,
    statistical_group = statistical_group
  )

table_S5B <- observed_summary |>
  mutate(
    cultivar = as.character(cultivar)
  ) |>
  left_join(
    model_summary,
    by = "cultivar"
  ) |>
  mutate(
    cultivar = factor(
      cultivar,
      levels = c(
        "Sylvaner",
        "Arvine",
        "Chasselas",
        "Pinot Noir",
        "Gamay"
      )
    )
  ) |>
  arrange(cultivar) |>
  transmute(
    Cultivar = as.character(cultivar),
    `Surface (ha)` = round(
      surface_ha,
      0
    ),
    `FD+` = fd_positive,
    `Observed FD+ ha-1` = sprintf(
      "%.2f",
      observed_fd_ha
    ),
    `Estimated FD+ ha-1 (95% CI)` = sprintf(
      "%.2f (95%% CI: %.2f-%.2f)",
      estimated_fd_ha,
      lower_CL,
      upper_CL
    ),
    `Statistical group` = statistical_group
  )

kable(
  table_S5B,
  caption = paste(
    "Table S5B. Observed and model-based FD-positive grapevine",
    "density among the five main grapevine cultivars included in",
    "the macro-scale study. Observed values correspond to vineyard",
    "surface, the total number of FD-positive grapevines and the",
    "observed FD-positive grapevine density. Estimated values are",
    "expressed as FD-positive grapevines per hectare with 95%",
    "confidence intervals. Different letters indicate significant",
    "differences based on Tukey-adjusted pairwise comparisons",
    "(α = 0.05)."
  )
)
```

```{r}
#| label: table-S5B-export

csv_file_S5B <- "tables/Table_S5B.csv"

write.csv(
  table_S5B,
  csv_file_S5B,
  row.names = FALSE,
  fileEncoding = "UTF-8"
)

cat(
  "CSV successfully created at:\n",
  normalizePath(
    csv_file_S5B,
    winslash = "\\",
    mustWork = TRUE
  )
)
```
